thesis2024

### 2025-03-27 thesis2024.tex
Creating frames article added to book.

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This is a book created from markdown and executable code.

See Knuth (1984) for additional discussion of literate programming.

@@

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

This research aims to deepen our understanding of the technological,
cultural, and social adaptations of hunter-gatherer groups during the
Upper Palaeolithic (UP) in southwestern Europe. By examining patterns of
resilience and innovation in raw material procurement and use, alongside
hunter-gatherer mobility, this study investigates how these behaviours
correlate with environmental changes and the evolving dynamics of social
territories. The focus is on the Iberian Peninsula, particularly
southwestern Iberia. Ultimately, this project will contribute to a
broader understanding of how past human groups lived and adapted
throughout the Upper Palaeolithic.

The UP (c.~40,000--10,000 BP) represents a key moment in human history,
marked by the expansion of modern humans into Europe, significant
cultural shifts, and climatic upheavals that transformed hunter-gatherer
lifestyles (Bar-Yosef 2002; Nuno Bicho et al. 2017a; Gamble et al. 2004;
Straus 1991, 1995; Sanchez Goñi and Harrison 2010b). These factors have
made the UP a central focus in prehistoric archaeology, particularly
regarding human adaptations and cultural transitions.

The Iberian Peninsula has played a critical role in this research. Its
cultural and geographical specificities offer valuable opportunities to
study cultural continuity and disruption through time within a single
site or region.

The impact of climate on human culture, resilience, and mobility is one
of the key research topics of the UP in Iberia (e.g., Nuno Bicho et al.
2017a; Nuno Bicho and Haws 2012; Cascalheira and Bicho 2013; Morin 2008;
Straus 1991; Bradtmöller et al. 2012; Schmidt et al. 2012). Patterns of
change and adaptation have been identified primarily through lithic
technological analyses, leading to the definition of the major
technocomplexes that characterise the UP in Iberia: Aurignacian,
Gravettian, Proto-Solutrean, Solutrean, and Magdalenian. These
technocomplexes exhibit distinct knapping strategies, technological
patterns, and raw material use, reflecting the technological
reorganisation of past human groups. Additionally, changes in territory
expansion, mobility, and social networks can be observed through
variations in technological choices and the presence of long-distance
raw materials (Bar-Yosef 2002; Zilhão 1997; Schmidt et al. 2012; Straus
1991; Fullola Pericot 1979).

The Iberian Peninsula has often been described as a refugium during the
coldest periods of the Late Pleistocene (Gómez and Lunt 2007a;
González-Sampériz et al. 2010; Hewitt 2000; Jochim 1987; Consuegra et
al. 2002; Jones 2013; Nuno Bicho, Cascalheira, and Gonçalves 2017),
providing milder conditions than northern Europe and supporting stable
human occupation throughout the UP. However, during abrupt climatic
events such as Heinrich Events and the Last Glacial Maximum, the region
experienced colder temperatures, reduced precipitation, and vegetation
shifts. Even within Iberia, north/south differences have been
identified: during abrupt climatic events, arboreal taxa such as oak
were replaced by infertile heathlands in northern Iberia or semi-desert
steppe vegetation in southern Iberia (Zapata et al. 2003; W. J. Fletcher
et al. 2010; W. Fletcher and Sanchez Goñi 2008; Naughton et al. 2009,
2007; Sanchez Goñi et al. 2000; Sánchez Goñi et al. 2008; K. Roucoux et
al. 2001; Boessenkool et al. 2001; Turon, Lezine, and Denèfle 2003; K.
H. Roucoux et al. 2005). This would have resulted in the expansion of
dry, risk-prone environments. Instead of being uniformly hospitable,
Iberia may have exhibited a ``leopard coat'' distribution of ecological
niches (Schmidt et al. 2012; Jennings et al. 2011). These niches likely
supported continuous human occupation, leading to the formation of
multicomponent sites and eco-cultural niches (Yravedra et al. 2016; Haws
2012; Cascalheira et al. 2017; Cascalheira and Bicho 2013; Banks et al.
2008; Fortea Pérez and Jordá Cerda 1975).

Some studies suggest a chronological correlation between climatic shifts
and the emergence of major technocomplexes, implying that cultural
reorganization may have been triggered by environmental changes that
altered landscapes and essential resources (e.g., Bradtmöller et al.
2012; Cascalheira and Bicho 2013; Nuno Bicho et al. 2017b). These
studies also indicate shifts in territory expansion/contraction,
settlement types, mobility, and intergroup contacts (Schmidt et al.
2012). While cultural changes are not necessarily deterministic
responses to climate fluctuations, it is evident that hunter-gatherers
adapted to their landscapes. Cultural, technological, and social
transformations provide key insights into how and why human groups
adapted at different times.

The study of raw materials, alongside lithic technological analyses, is
a powerful tool for identifying patterns of change and adaptation,
particularly in landscape use and mobility. Raw materials found at
archaeological sites reflect human decision-making and technological
organisation (Bamforth and Bleed 1997). These patterns provide crucial
insights into how hunter-gatherers interacted with their environments
and adapted to shifting conditions (Féblot-Augustins 1993; Kuhn 1995;
Mellars 1996). Variability in raw material use at a given site may
result from factors such as mobility and procurement strategies
influenced by climate (Ambrose and Lorenz 1990; Binford and Stone 1985;
Kuhn 1991, 2004; Gould and Saggers 1985; McCall 2007), settlement
patterns (Kuhn 2004; Surovell 2009; Grove et al. 2023), or exchange
networks and social interactions
(\textbf{akermanWeaponsWunanProduction2002?}; Wurz 1999; Whallon 2006;
Robin Torrence 1986).

The study of raw materials from UP sites in Iberia, particularly
chert---a key raw material widely used for tool production---has
provided valuable insights into hunter-gatherer procurement strategies,
mobility patterns, and social networks.

A clear example is the identification of long-distance contacts between
regions in Iberia, suggesting complex mobility and procurement patterns
throughout the UP (e.g., Hermida, Rellán, and Rodríguez 2016; Corchón
Rodríguez, Martínez, and Tarriño 2016; Ortega 2003; Sánchez de la Torre
et al. 2023). The Côa Valley in northeastern Portugal is a particularly
emblematic case, as non-local chert is present throughout the UP
sequence, with potential sources located 150--250 km away in both
southwestern (central Portugal) and southeastern (central Spain)
directions. These patterns have been interpreted as evidence of exchange
networks facilitated by seasonal movements and social interactions
between groups across Iberia (Aubry et al. 2022, 2004).

Compared to other Iberian regions, where raw material studies are more
common, knowledge of UP hunter-gatherer lifeways, organisation, and
social networks in southwestern Iberia has primarily been derived from
lithic technology and faunal analyses. Much of this research has focused
on Vale Boi, the only well-preserved UP site in the region. This
presents a unique opportunity to investigate hunter-gatherer adaptations
in a key area of human occupation during this period.

Vale Boi, located in southwestern Iberia, is a multi-component site
spanning several UP technocomplexes, from the Gravettian (c.~32 ka cal
BP) to the Magdalenian {[}c.~15 ka cal BP; Nuno Bicho et al. (2007);
Nuno Bicho, Cascalheira, and Marreiros (2012); Nuno Bicho et al. (2013);
Cascalheira et al. (2017); Cascalheira and Bicho (2013); Manne et al.
(2012); Manne and Bicho (2011){]}. The site, considered an eco-cultural
niche, provided stable climatic conditions even during colder periods,
supporting continuous residential occupations that have been
consistently dated (Nuno Bicho et al. 2013; Cascalheira et al. 2017;
Belmiro et al. 2021). Extensive lithic assemblages from Vale Boi have
undergone technological analysis, revealing patterns of technological
change and resilience (Belmiro et al. 2021; Cascalheira et al. 2013;
Gibaja Bao and Ferreira Bicho 2013; Horta, Cascalheira, and Bicho 2019;
Marreiros et al. 2015). The site has also been identified as a centre
for knowledge transfer, particularly during the Last Glacial Maximum
(\textbf{cascalheiraINFLUENCIAMEDITERRANICANAS2013?}; Cascalheira 2019).

Despite extensive technological studies on Vale Boi's lithic
assemblages, macroscopic, microscopic, and geochemical analyses of chert
remain limited (Verissimo 2004; Pereira, Bicho, et al. 2016). This gap
hinders research into raw material procurement and use, mobility, and
connections between southwestern Iberia and other regions.

Consequently, Vale Boi presents an exceptional opportunity to advance
understanding of UP hunter-gatherer adaptations, mobility, and
technological and social organisation in Iberia.

Focusing on the lithic assemblages from Vale Boi, this research aims to
investigate chert procurement, use, and mobility patterns throughout the
UP of southwestern Iberia, as well as their relationship with the
techno-cultural dynamics of each technocomplex and climatic changes.

To achieve this, I will address the following research questions:

\begin{itemize}
\item
  What types of chert were exploited by hunter-gatherer groups in
  southwestern Portugal, and how were these materials introduced to the
  site?
\item
  How frequently did these groups move across the landscape, and can
  long-distance movements or social contacts be identified through the
  presence of non-local cherts?
\item
  How were different chert types managed by hunter-gatherer groups, and
  what impact did mobility have on their technological organisation?
\item
  How did technological and social organisation evolve over
  approximately 10,000 years of the UP in southwestern Portugal, and how
  might climatic changes have influenced these adaptations?
\end{itemize}

To address these questions, this research employs a multi-method
approach, integrating geological surveys with macroscopic, petrographic,
mineralogical, and geochemical analyses, alongside technological studies
of chert lithic artefacts.

The research is structured into three peer-reviewed articles (Chapters
2, 3, and 5) and one additional chapter, each addressing different
aspects of chert procurement and usage. A concluding discussion
integrates the findings and examines their broader implications for
understanding human behaviour and adaptation over time.

\begin{itemize}
\item
  Chapter 2, titled ``Creating Frames of Reference for Chert
  Exploitation during the Late Pleistocene in Southwesternmost Iberia''
  (Belmiro, Terradas, and Cascalheira 2023) offers a geological
  characterisation of chert sources in the Algarve region, employing
  macroscopic and petrographic (thin section analysis) methods. It
  presents a reference collection and database of local and regional
  chert resources, which is crucial for distinguishing between local and
  non-local cherts when analysing an archaeological assemblage.
\item
  Chapter 3, titled ``Within and Beyond: Chert Procurement Patterns
  During the Upper Palaeolithic in Southwesternmost Iberia'' (Belmiro et
  al. 2025) focuses on the analysis of chert artefacts from the UP
  sequence at the archaeological site of Vale Boi, using both
  macroscopic and petrographic (thin section analysis) methods. These
  artefacts are compared with geological reference collections. The
  chapter offers new insights into the presence of local and non-local
  cherts across the various UP occupations, identifies the probable
  sources of non-local cherts, and explores changes in the distribution
  of chert types over time.
\item
  Chapter 4, titled ``Within Chert: A Multi-Technique Mineral and
  Geochemical Approach to the Study of Chert in Southwestern Iberia,''
  builds on the previous chapters by applying various techniques to
  analyse geological and archaeological chert samples. It provides data
  on the mineral and chemical composition of cherts through X-Ray
  Diffraction (XRD), Scanning Electron Microscopy and Energy Dispersive
  X-Ray Spectroscopy (SEM-EDS), and portable X-Ray Fluorescence (pXRF).
  These methods allow for a more comprehensive characterisation of the
  samples and enable the use of chemical composition to refine the
  source attributions of archaeological cherts.
\item
  Chapter 5, titled ``From Stone to Tool: How Raw Materials Influenced
  Upper Palaeolithic Technology at Vale Boi, Southwestern Iberia,''
  combines the previously obtained data on chert types (particularly
  regarding the distinction between local and non-local cherts) with
  technological data. It explores various patterns of chert use and
  management.
\item
  Chapter 6 synthesises the results and contributions of the previous
  chapters and discusses the broader implications of these patterns for
  understanding human behaviour.
\end{itemize}

Together, these papers and chapters present the first comprehensive
study of chert raw materials associated with technology from UP
occupations in southwestern Iberia and contribute to the ongoing
discussion on past human adaptations.

The creation of an online database freely accessible to the scientific
community provides the first online reference collection for cherts in
the region. This is crucial for the development of accessible and
transparent scientific production, as it offers a valuable point of
reference for future studies on chert across various prehistoric
periods, not only in the region but also throughout the Iberian
Peninsula.

@@

\textbf{Authors}: Joana Belmiro, Xavier Terradas, João Cascalheira

Southwestern Iberia has played a key role in characterizing Late
Pleistocene human ecodynamics. Among other aspects of human behavior,
chert procurement and management studies in this region have received
increasing attention in the past two decades, especially focusing on the
sites showing repeated human occupation, such as the case of Vale Boi
(Southern Portugal). However, these studies have been very limited in
their geographical scope, and mostly focused on brief macroscopic
descriptions of the raw materials. To further our knowledge of the
relationship between regional availability of raw materials and its
impact on human adaptations and mobility, a more detailed approach to
characterizing geological sources is needed. This paper characterizes
chert raw materials location, diversity, and availability in a
geologically well-defined region of southern Portugal - the Algarve.
Through macroscopic and petrographic approaches, we provide a detailed
characterization of geological chert sources to build a frame of
reference for chert exploitation in the region. Our results show that
there are four main chert formations in Algarve, and that despite the
within-source variability, sufficient differences at macroscopic and
petrographic levels are present to allow clear source attribution. These
results provide a baseline for raw material studies in archaeological
assemblages across southwestern Iberia, that will be essential to
further characterize the dynamics of human behavior in some of the most
important eco-cultural niches.

Southern and Western Iberia have often been considered key areas to
understanding techno-cultural transitions from the Middle Paleolithic to
the end of the Upper Paleolithic. As territories located at the western
tip of the European continent and with a generally stable climate even
during the coldest periods punctuating the Late Pleistocene, these have
been regarded as some of the most significant glacial refugia in Europe
(Carvalho et al. 2022; Gómez and Lunt 2007b; González-Sampériz et al.
2010; Hewitt 2000). For this reason, Southwestern Iberia has frequently
been at the center of some of the most debated topics regarding Late
Pleistocene human adaptations (Cascalheira et al. 2017; Finlayson et al.
2006; Zilhão et al. 2017). A particularly good example is the region's
role in possibly being one of the last territories to be occupied by
Neanderthal populations right before their complete disappearance
(Finlayson et al. 2006; Zilhão et al. 2017).

Neanderthals occupied the European continent for more than 300.000 years
and are thought to have disappeared while modern humans arrived on the
territory (Finlayson et al. 2006; Jennings et al. 2011; Mellars 2004;
Rogers, Bohlender, and Huff 2017; Trinkaus and Howells 1979). The last
territory where Neanderthal populations seemed to exist was Southern
Iberia, around c.~37 thousand years ago (Zilhão et al. 2017), or perhaps
even earlier (Carvalho et al. 2022; Finlayson et al. 2006; Tzedakis et
al. 2007). This region is key to understanding how Neanderthals survived
until such a later chronology, the degree and types of interaction those
populations may have had with modern humans (Finlayson et al. 2006;
Zilhão et al. 2017), and how and why they eventually went extinct (Dalen
et al. 2012; Melchionna et al. 2018). Another example is Southwestern
Iberia's importance in the discussion of Upper Paleolithic
technocomplexes transitions. Previous studies have discussed the
territory's potential as a refugium during cold and harsh climatic
conditions (Gómez and Lunt 2007b; Jennings et al. 2011;
Rodríguez-Sánchez et al. 2010). The Heinrich Event 2 (HE 2) at the onset
of the Last Glacial Maximum (LGM), for example, is a period marked by
important social and technological transformations. This climatic event
was characterized by abrupt and drastic climatic changes that impacted
human behavior all across westernmost Europe (Gamble et al. 2004;
Sanchez Goñi and Harrison 2010a). The identification of a
Proto-Solutrean phase in central and southern Portugal with a very
distinct index fossil (the Vale Comprido point), and its direct
association with the HE 2 (Cascalheira and Bicho 2013), put these
regions amongst some of the most important case studies of how
environmental dynamics have affected human adaptations during the last
glacial. Other studies have expanded upon this notion of climatic
refugia during harsh climatic events, to understand the Iberian
Peninsula as a long-term eco-cultural refugia (Cascalheira et al. 2017).
Using this framework, this territory would consist of several ecological
niches, consistently used through time, possibly due to the stability in
the richness and variety of resources. This continuous use would then
create long-term regional adaptive structures, which when correlated
with the ecological niches, Cascalheira et al. (2017) have referred to
as eco-cultural niches. In fact, it seems that a large number of caves
and rockshelters in Iberia are multi-layered, giving validity to the
aforementioned framework (Schmidt et al. 2012). These eco-cultural
niches provide an exceptional opportunity to understand long-term
dynamics regarding biotic and abiotic resource exploitation since they
can provide details on how human populations maintained or changed their
adaptive systems when facing environmental changes, and cultural and
social transformations or constraints. One of such possible Late
Pleistocene eco-cultural niches to which this theoretical framework has
been previously applied is the archaeological site of Vale Boi. This
multi-component site is located in westernmost southern Iberia, in a
region currently known as the Algarve, and comprises one of the most
complete Upper Paleolithic chronocultural sequences of southern Iberia
(Nuno Bicho et al. 2013; Cascalheira and Bicho 2013; Cascalheira et al.
2017).

Several cross-scale complex interactions have been identified,
displaying resilient behaviors throughout the Upper Paleolithic
maintained by their eco-cultural niche, but also adaptation behaviors
motivated by niche diversity, social networks, and climatic changes
(Cascalheira et al. 2017). Some of these resilient elements are, for
example, the continuous use of strategies like grease rendering and
selective hunting patterns, site function, certain lithic technology
patterns, and the functional specialization of lithic raw materials. The
maintenance of social networks through the identification of possible
long-distance lithic raw materials has also been suggested (Pereira,
Bicho, et al. 2016). The identification of these patterns has been
reliant on the great amount of studies that have originated from the
archaeological site of Vale Boi (i.e., Pereira, Bicho, et al. 2016; Nuno
Bicho et al. 2017b; Nuno Bicho and Haws 2012; Manne et al. 2012). A
large portion of these studies has focused on lithic technology (Belmiro
2020; Cascalheira 2010; Marreiros et al. 2015), and unlike other regions
of Portugal (Aubry et al. 2022, 2004; Aubry and Igreja 2009; Costa et
al. 2022; Jordão and Pimentel 2022; Pereira et al. 2021, 2022),
archaeological studies focusing on raw materials, and especially chert,
have been more scarce, focusing mostly on brief macroscopic results or
the differentiation between possibly local and non-local sorts of raw
materials. Despite the scarceness of these studies, chert played a
significant role in early and later prehistory at Vale Boi (Belmiro
2020; Cascalheira 2010; Marreiros 2009), and across most of Prehistory
in Southern Portugal (Nuno Bicho et al. 2003; Zambujo 1998). As an
essential part of late Pleistocene hunter-gatherer adaptations, lithic
raw materials have the potential to provide insights into the adaptive
strategies of those populations (Féblot-Augustins 1993; Kuhn 1995;
Mellars 1996), mostly regarding land-use, technological organization,
but also cultural and social interactions. In fact, the selection and
procurement of raw material has been suggested as a key stage for the
technological organization of hunter gatherer groups (Bamforth and Bleed
1997). Thus, changes in the frequencies of raw materials within the
archaeological records may provide evidence for changes in that
organization, or even resilience of specific choices, all of which can
reflect culturally transmitted preferences within a group of hunter
gatherers (Brown 1999). Several models, both formal and informal, have
shown that changes of raw material in the archaeological record may
reflect different procurement strategies (Binford 1979; Binford and
Stone 1985; Gould 1985; Gould and Saggers 1985), either related to
mobility strategies (Ambrose and Lorenz 1990; Kuhn 1991, 2004; McCall
2007) or changes in the availability of raw materials, possibly related
to environmental change (P. Brantingham 2003; Brown 1999). Raw materials
are also closely linked to the tools produced with them -- the costs of
manufacturing technology can be related to the availability of raw
materials in the landscape (Bousman 1993; Mackay and Marwick 2011; Parry
and Kelly 1987; R. Torrence 1983), or the characteristics of the raw
materials which might relate to functionality (Mackay 2008; Minichillo
2006; Stout 2002), through the choice of raw materials for specific
technologies (Brown 1999). Furthermore, changes in the raw materials can
also be related to the establishment of social networks (Whallon 2006)
and the horizontal transmission of preferences through trade (Brown
1999). In order to explore the aforementioned theories and models in an
archaeological site, especially when trying to understand land-use and
the adaptive strategies of an eco-cultural niche like what is proposed
for southwesternmost Iberia, it is necessary to know the landscape and
the raw material sources available in the territory. To distinguish
between local and non-local raw materials, first, it is necessary to
understand the characteristics of the local sources and establish a
comparative database that can be used for the analysis of an assemblage.
This is true for most of the informal models which try to understand why
raw materials change in the archaeological record. Formal models
focusing on raw material use posit that it is necessary to have detailed
sourcing data in order to adequately apply the models to a region or an
archaeological assemblage (Pop 2015; Surovell 2009). In fact, the
creation of such a database is a starting point for most raw
material-focused studies, independent of the geography or studied
chronology (Bustillo et al. 2009; Driscoll, Burke, and Warren 2016;
Ekshtain and Zaidner 2022; Luedtke 1992). In western Europe, there have
been substantial efforts in creating comprehensive knowledge about
chert-bearing formations, which resulted in important lithotheques and
databases (Sánchez et al. 2014; Paul Fernandes et al. 2013; Ortega and
Terradas 2014), and multidisciplinary studies of raw material use
throughout prehistory (Aubry et al. 2022; Delagnes et al. 2006; P.
Fernandes et al. 2012; Rodríguez, Rodríguez, and Pelegrin 2011; Terradas
and Ortega 2017; Turq et al. 2017), as early as the 1980s.

In southern Portugal, previous efforts have been made to create such a
database. The work of Verissimo (2004), albeit focusing on the
occurrence of chert solely in western Algarve, provided the initial
basis for comparative studies with the assemblages from some Late
Pleistocene sites based on macroscopic approaches. The creation of
LusoLit, a lithotheque currently hosted at the University of Algarve
(Pereira, Farias, and Paixão 2016), and the collection of samples from
several outcrops in the region provided a new leap in the study of chert
in the region. A few geological studies have also contributed to
understand the availability and characteristics of chert in southern
Portugal (Ribeiro 2005; Rogério Rocha 1976). Nevertheless, these studies
are often unpublished or comprehend answers to geological questions,
which hamper the comparative use of the data with archaeological
assemblages. A large portion of chert-bearing outcrops in southern
Portugal remains unstudied, both by archaeologists and geologists. As
such, further analyses of the overall variability, location, and
availability of chert nodules in the Algarve are necessary to start
testing behavioral models of land-use and abiotic resource exploitation.
Given the potential for an in-depth raw material study at a possible
Late Pleistocene eco-cultural niche such as Vale Boi---also due to the
ubiquitous presence of chert throughout the stratigraphy, as one of the
main raw materials used for lithic technology in the site (Belmiro 2020;
Cascalheira 2010; Marreiros 2009) ---in this study we explore the
location, diversity, and availability of chert raw materials in the
southernmost region of Portugal -- the Algarve. Our main goal is to
establish a reference for chert raw materials in an understudied region
in regards to lithic materials, that can be used for future studies
addressing chert exploitation in the Algarve and elsewhere. This
includes the development of a methodological approach adapted to the
study area, by testing the potential of macroscopic and petrographic
approaches for the characterization of regional cherts. The data
presented here is also fully integrated into an online lithotheque
(LusoLit) that is now freely available and can be built upon and
improved as new data becomes available.

\section{Geological setting and chert
groups}\label{geological-setting-and-chert-groups}

\subsection{Geological setting}\label{geological-setting}

The Algarve is the southernmost region of Portugal, framed north by the
Alentejo region and east by Spain. To the west and south, it is bordered
by the Atlantic ocean. It extends for \textasciitilde130 km E-W and
\textasciitilde50 km N-S and is characterized by a variety of geomorphic
sub-regions and geological units, that make this region a complex
territory (@fig-frames-geomap). On the north sector of the Algarve, the
Serra Algarvia is characterized by a mountainous range with a dense
hydrographic network, which separates the Algarve from Alentejo. On the
south sector, the Litoral is characterized by a flatter, long strip of
land, that extends through all of the coastal strip of the Algarve. The
Barrocal is nested between the other sub-regions and has a more moderate
relief, characterized by carbonated Jurassic formations and important
sub-terranean water circulation (Gago 2007).

\includegraphics{figures/creating-frames/map_layout.png}

\caption{\label{fig-frames-geomap}Geological map of the Algarve region.
The map represents the several geological layers and formations, as well
geomorphic sub-regions. The geological data was obtained through the
vectorization of the geological raster maps (Carta Geológica de
Portugal, 1:500 000 scale) made available by LNEG (Laboratório Nacional
de Energia e Geologia) in
https://geoportal.lneg.pt/pt/dados\_abertos/cartografia\_geologica/cgp500k/folhasul/.
The inset map was created from a 30×30‐m pixel Digital Elevation Model
(DEM) obtained from the Shuttle Radar Topography Mission
(https://www2.jpl.nasa.gov/srtm/dataprod.htm).}

Geologically, the Algarve is composed of two main geological units: the
South Portuguese Zone (SPZ) and the Algarve basin. The SPZ is located in
the north sector of the Algarve, extending up to Alentejo (Paulo
Fernandes et al. 2012). Its main lithologies are schist, greywacke and
quartzite (Paulo Fernandes et al. 2012; Pereira, Bicho, et al. 2016).
The SPZ is overlain unconformably by the Mesozoic sedimentary rocks of
the Algarve basin (Paulo Fernandes et al. 2012). The basin corresponds
to the Mesozoic-Cenozoic sediments that outcrop south of the Algarve,
from the westernmost to the easternmost point of the region, and it is
associated with the opening of the central Atlantic Ocean and with the
eventual oceanic crust formation in the western part of the Tethys sea,
between the Algarve and North Africa (Terrinha et al. 2006). Mesozoic
sedimentation of the basin started in the Triassic and continued
thereon. In the Lower Jurassic (Lower Pliensbachian, also regionally
known as Carixian) the basin was divided into two sub-basins -- western
and eastern sub-basins (Rogério Rocha 1976; Terrinha et al. 2006). The
existence of the two sub-basins and the expansion and retraction of the
sea created a variety of sedimentation environments, such as external
and internal platforms, continental, hemipelagic and deep marine
(Terrinha et al. 2006), as well as moments of sedimentation hiatus. This
variability in deposition environments created a variety of sedimentary
facies, with moments of more or less homogeneity throughout this period.
For example, during the Lower Pliensbachian, in the Lower Jurassic, the
sediments in the western sub-basin can be described as marine of
external platform, while the sediments of the eastern sub-basin can be
described as marine of internal platform. During the Upper Jurassic
however, the basin is marked by a moment of prominent lithofacies
variation, followed by a moment of uniformity in both sub-basins
(Terrinha et al. 2006). Understanding the Algarve basin is key for raw
material studies in the Algarve, especially when studying chert, since
it is in the basin, more specifically in the Jurassic sediments, where
chert primarily outcrops in the region.

\subsection{Chert outcrops}\label{chert-outcrops}

Within the Algarve area, the presence of chert may be associated with
carbonates in limestone and dolomite formations. This is explained by
characteristics (such as the presence of water or specific temperatures
and pH) that are ideal for both the formation of limestone and the
precipitation of silica (Luedtke 1992). The pelagic and marine
environments of the Algarve basin during the Jurassic gathered those
such ideal characteristics: as shown by Ribeiro (Ribeiro 2005), the
cherts from the Lower Jurassic are the result of early diagenic
silicification of carbonate sediments. The existence of two basins with
different sedimentation environments also shows potential for the
existence of different types of chert throughout the basin and their
differentiation. For example, during sedimentation, skeletal grains of
fossils may be preserved. Many of these fossils are restricted to
specific environments and time intervals (Flügel 2010), which may allow
the identification of chert outcrops through the fossil content. The
basin and sub-basins thus show potential for the existence of different
geological formations with different chert types and their study.
Previous works, both geological and archaeological, confirm that chert
is present in the Algarve in the Jurassic limestone or dolomitic
limestone layers of the Algarve basin. This means that chert outcrops
can be identified in the central/south sector of the region, from west
to east. Variability in chert availability, as well as chert types, is
expected, considering that during the sedimentation process, the Algarve
basin was already sub-divided and in constant environmental change. Due
to this, several formations with chert nodules can be identified in the
Algarve, attributed to different sub-periods of the Jurassic. In the
western sector of the Algarve, chert can be found in the Lower Jurassic
(Carixian) formations, in limestone or dolomitic limestone layers
(Rogério Rocha 1976; R. Rocha et al. 1979), often visible in areas where
the layers are exposed, such as beach-generated cliffs and associated
deposits, like Cabo de S. Vicente (CSV) and Praia do Belixe (PBX). These
formations with chert nodules are also visible inland, albeit more
scarcely, as is the case of the small outcrop named Ferrel (FER), 3 km
from the current coastline. Lower Jurassic chert-bearing formations are
barely existent in the center/east sector of the Algarve, with one
single formation with micro-nodules identified in geological works
(Oliveira 1992). Middle Jurassic geological layers with chert nodules
are only found in the center/east sector of the Algarve in a geological
formation called the Malhão formation. The formation can be described as
carbonated, from a marine sedimentation environment. Chert in this
formation has been identified in two distinct layers: conglomerates with
micritic limestone intercalations with chert beds and nodules,
characterized by the presence of sponge spicules and radiolarians
(Manuppella et al., 1987); microcrystalline limestones with chert
nodules characterized by the presence of silicified malacofauna and
silicified corals (Manuppella et al. 2007). Finally, Upper Jurassic
sediments with chert nodules have also been mostly identified in the
center/east sector of the Algarve, attributed to the Jordana formation.
This formation is characterized by dark-gray limestones, with frequent
secondary silicifications with abundant fossil fragments (Manuppella et
al. 1987; Manuppella et al. 2007; R. Rocha et al. 1989). Upper Jurassic
sediments with chert nodules in western Algarve (Kimmeridgian
formations) have only been identified in one area, between Ponta da
Atalia (PtA) and Praia da Mareta (MAR) (R. Rocha et al. 1979). Given the
differences of the cherts and formations between the western and eastern
sectors of the Algarve already established in previous works, this
division will be followed in the present study.
